Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-11-24 concurrency/CPP 2024/paper.tex
WIP

\lstdefinestyle{myStyle}{
	backgroundcolor=\color{white}, 
	keywordstyle=\bfseries\color{green!40!black},
	commentstyle=\itshape\color{violet!40!black},
	stringstyle=\color{codepurple},
	basicstyle=\ttfamily\footnotesize,
	breakatwhitespace=false,         
	breaklines=true,                 
	keepspaces=true,                 
	numbers=left,       
	numbersep=5pt,                  
	showspaces=false,                
	showstringspaces=false,
	showtabs=false,                  
	tabsize=2,
}

@@

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
		%		\label{traverse_lock_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{insert.ml} 
	\end{subfigure}
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		%		\label{traverse_lock_b}
	\caption{The lock-coupling search structure template}
	\label{template-ex}
\end{figure}
Figure~\ref{template-ex} shows an example search structure template. The core of the template is the \lstinline{traverse} function, which uses a specific concurrency control mechanism to travel through a data structure in search of the requested key. The mechanism in this example is \emph{lock coupling}, where we acquire the lock on the next node before releasing the lock on the current node. The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine. Once the appropriate node for the key has been found, the template returns it to a top-level function such as \lstinline{insert} that calls out to the data structure to perform the actual insertion on the node. Thus, the \lstinline{traverse} and \lstinline{insert} functions can be verified without knowing anything about the target data structure other than its synchronization mechanism, as long as the data structure implements \lstinline{findNext} and \lstinline{insertOp} operations with the required semantics. The concurrency functions in the template and the sequential \lstinline{findNext} and \lstinline{insertOp} functions provided by the data structure combine into a fully operational concurrent data structure. Krishna et al.~specify templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures (i.e., those in which a node may be reachable along multiple paths), but many common search structures do not require flow-style reasoning; in this paper, we present a version of templates that is independent of flow interfaces. %is this a contribution in itself?
%what more do we want to say here?

@@

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
$$
 {\color{specblue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(s, v)\  {\color{specblue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
\begin{mathpar}
	{\color{specblue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ } 
	 \vspace{-0.8em}  \\ \texttt{push}(s, v)\   \vspace{-0.8em}  \\ 
	{\color{specblue} \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle }
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

The core of a search structure template is a \lstinline{traverse} function that navigates a data structure using the chosen synchronization mechanism. A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well. The defining feature of all these functions is that they abstract away from the data structure being traversed: they operate on a generic \lstinline{node} type and take data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters. We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
\begin{lstlisting}[style=myStyle,language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, numbers=none, xleftmargin=0.5em]
	typedef struct node_t {node *t; lock_t *lock;} node_t;
\end{lstlisting}
There are two things worth noting about this data structure. First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding locks on as few nodes as possible to allow other parts of the data structure to be modified concurrently. Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as 
\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *left, *right;
	} node;
\end{lstlisting}
while a linked list might use
\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *next;
	} node;
\end{lstlisting}
This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component. Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns, and then prove (assuming appropriate specifications for data-structure-specific functions) that they correctly implement the data structure operations.

@@

\begin{mathpar}
	{\color{specblue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
	 \vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
\\
	{\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m) \right\rangle}\ 
	 \vspace{-0.8em} \\ \texttt{release}(n\texttt{->lock})\  \vspace{-0.8em} \\ {\color{specblue}\left\langle \nodeboxrep(n) \ |\ \treerep\ (m)\right\rangle}
\end{mathpar}

@@

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ast \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys). Then $\treerep\ (m)$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \nodeboxrep(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}
\begin{figure*}[!ht]
	$\color{specblue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{specblue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast  \infp (r) \ast \infp (n)
\end{align*}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

>>>>

\lstdefinestyle{myStyle}{
	backgroundcolor=\color{white},
	belowcaptionskip=1\baselineskip,
	frame=none,
	columns=[c]fixed,
	keywordstyle=\bfseries\color{green!40!black},
	commentstyle=\itshape\color{violet!40!black},
	stringstyle=\color{codepurple},
	basicstyle=\ttfamily\footnotesize,
	breakatwhitespace=false,         
	breaklines=true,                 
	keepspaces=true,                 
	numbers=left,       
	numbersep=5pt,                  
	showspaces=false,                
	showstringspaces=false,
	showtabs=false,                  
	tabsize=2,
}

@@

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
		%		\label{traverse_lock_a}	
	\end{subfigure}
	\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{insert.ml} 
	\end{subfigure}
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		%		\label{traverse_lock_b}
	\caption{The lock-coupling search structure template}
	\label{template-ex}
\end{figure}

Figure~\ref{template-ex} shows an example search structure template. The core of the template is the \lstinline{traverse} function, which uses a specific concurrency control mechanism to travel through a data structure in search of the requested key. The mechanism in this example is \emph{lock coupling}, where we acquire the lock on the next node before releasing the lock on the current node. The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine. Once the appropriate node for the key has been found, the template returns it to a top-level function such as \lstinline{insert} that calls out to the data structure to perform the actual insertion on the node. Thus, the \lstinline{traverse} and \lstinline{insert} functions can be verified without knowing anything about the target data structure other than its synchronization mechanism, as long as the data structure implements \lstinline{findNext} and \lstinline{insertOp} operations with the required semantics. The concurrency functions in the template and the sequential \lstinline{findNext} and \lstinline{insertOp} functions provided by the data structure combine into a fully operational concurrent data structure. Krishna et al.~specify templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures (i.e., those in which a node may be reachable along multiple paths), but many common search structures do not require flow-style reasoning; in this paper, we present a version of templates that is independent of flow interfaces. %is this a contribution in itself?
%what more do we want to say here?

@@

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as 
$$
 {\color{specblue} \{\mathsf{stack}\ s\ p\} } \ \texttt{push}(p, v)\  {\color{specblue} \{\mathsf{stack}\ (v :: s)\ p\}}$$  Its concurrent counterpart could be specified as 
\begin{mathpar}
	{\color{specblue} \fforall s.\ \left\langle \mathsf{is\_stack}_g\ s\ |\ \mathsf{stack}_g\ vs\right\rangle\ } 
	 \vspace{-0.8em}  \\ \texttt{push}(s, v)\   \vspace{-0.8em}  \\ 
	{\color{specblue} \left\langle \mathsf{is\_stack}_g\ s\ |\ \mathsf{stack}_g\ (v::vs)\right\rangle }
\end{mathpar}
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $vs$ to $v::vs$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

The core of a search structure template is a \lstinline{traverse} function that navigates a data structure using the chosen synchronization mechanism. A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well. The defining feature of all these functions is that they abstract away from the data structure being traversed: they operate on a generic \lstinline{node} type and take data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters. We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
\begin{lstlisting}[style=myStyle,language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, numbers=none, xleftmargin=0.5em]
	typedef struct node_t {
		node *t; 
		lock_t *lock;
	} node_t;
\end{lstlisting}
There are two things worth noting about this data structure. First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding locks on as few nodes as possible to allow other parts of the data structure to be modified concurrently. Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as 
\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *left, *right;
	} node;
\end{lstlisting}
while a linked list might use
\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *next;
	} node;
\end{lstlisting}
This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component. Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns, and then prove (assuming appropriate specifications for data-structure-specific functions) that they correctly implement the data structure operations.

@@

\begin{minipage}{.2\textwidth}
	\centering
\begin{mathpar}
	{\color{specblue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
	\vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
\end{mathpar}
\end{minipage}
\begin{minipage}{.28\textwidth}
		\centering
\begin{mathpar}
	{\color{specblue}\left\langle \nodeboxrep \left(n\right) \ |\ \treerep\ (m) \right\rangle}\ 
	\vspace{-0.8em} \\ \texttt{acquire}\left(n\texttt{->lock}\right)\  \vspace{-0.8em}  \\ {\color{specblue}\left\langle \nodeboxrep(n) \ast \mathsf{R}(n) \ |\ \treerep\ (m)\right\rangle}
\end{mathpar}
\end{minipage} \\

@@

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ast \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys). Then $\treerep\ (m)$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}, \texttt{k}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \nodeboxrep(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}

\begin{figure*}[!ht]
	$\color{specblue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{specblue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{align*} &\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\  & \ \ \ \ \ \ \ \ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast  \infp (r) \ast \infp (n)
\end{align*}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.
